---
title: "Labor Market Analysis"
author: "y_bib"
format:
  pdf:
    include-in-header: 
      text: |
        \usepackage{float}
        \usepackage{makecell}
        
editor: visual
---

# Labor Market Canada

**Labor market survey results** provided by Statistics Canada contain information on employment status, type of employment, gender, marital status, and whether children are present at home. For employed individuals, data on wages, hours worked, overtime, the type and field of occupation are included and etc. For part-time workers, the survey additionally explores the reasons for part-time employment. For unemployed individuals, the survey provides insights into the reasons for not working and related factors.

**Goals of the Analysis:**

-   For employed people: provide a description and test the statistical significance of wage differences based on gender, immigrant status, age, and other relevant variables.

-   Study which factors are most crucial in determining wage levels: education, place of living, occupation type and field, gender, age and etc.

-   Examine differences between women and men, immigrants and non-immigrants in terms of overtime work, part-time employment, and unemployment rates.

# Summary of results

```{r}
#| message: false
#| warning: false
#| include: false
#loading libraries
library(plyr)
library(dplyr)
library(ggplot2)
library(ggpmisc)
library(patchwork)
library(descr)
library(stringr) # for formatting input data - working with strings
library(pdftools) # reading pdf file
# library(lubridate) handling datas ant time - did not use it
library(dbscan) # for clustering
library(tidyr)
library(pivottabler) # for pivot tables
library(gt) # tables


#reading date from Stat Ca
labor<-read.table(file='data/pub0325.txt',header=TRUE,sep="\t",stringsAsFactors = TRUE)


names(labor) <- tolower(names(labor))

#indicating columns as factors
names<-as.vector(colnames(labor),mode="character")
names_factor<-setdiff(names,c("rec_num","finalwt","paidot","unpaidot",
        "xtrahrs","wksaway","uhrsmain","ahrsmain",
        "utothrs","atothrs","hrsaway","tenure","prevten","hrlyearn"))

labor[names_factor] <- lapply(labor[names_factor], as.factor)

labor$immig<-revalue(labor$immig, c("1" = "Immig,not more 10y", "2" = "Immig,more 10y", "3"="Non-immigrant"))
labor$gender <- revalue(labor$gender, c("1" = "Male", "2" = "Female"))
labor$ftptmain<-revalue(labor$ftptmain,c("1"="Full-time","2"="Part-time"))
labor$lfsstat<-revalue(labor$lfsstat,c("1"="Employed, at work",'2'="Employed, absent from work","3"="Unemployed","4"="Not in labour force"))


age_12_labels <- c(
  "1" = "15 to 19",
  "2" = "20 to 24",
  "3" = "25 to 29",
  "4" = "30 to 34",
  "5" = "35 to 39",
  "6" = "40 to 44",
  "7" = "45 to 49",
  "8" = "50 to 54",
  "9" = "55 to 59",
  "10" = "60 to 64",
  "11" = "65 to 69",
  "12" = "70+"
)

# Reasons for part time. whypt field
whypt_labels <- c(
  "0" = "Other reasons",
  "1" = "Own illness or disability",
  "2" = "Caring for children",
  "3" = "Other personal or family responsibilities",
  "4" = "Going to school",
  "5" = "Personal preference",
  "6" = "Business cond. or could not find full-time work, looked for it in last month",
"7"=	"Business cond. or could not find full-time work, did not look for it in last month")


```

# Descriptive Statistics

## Wages depending on gender

Mean and median of hourly rate vs gender

```{r}
#| echo: false
#| message: false
#| warning: false


tapply(0.01*labor$hrlyearn, labor$gender , mean, na.rm = TRUE)

tapply(0.01*labor$hrlyearn, labor$gender , mean, na.rm = TRUE)

```

Mean and median wages vs gender

```{r}
#| echo: false
#| message: false
#| warning: false


#hour rate* usual hours=income  vs gender

tapply(0.01*labor$hrlyearn*labor$utothrs, labor$gender , mean, na.rm = TRUE)
#boxplot(hrlyearn ~ gender ,data=labor, ylim=c(0,15000))

#hour rate* usual hours=income  vs gender

tapply(0.01*labor$hrlyearn*labor$utothrs, labor$gender , median, na.rm = TRUE)
#boxplot(hrlyearn ~ gender ,data=labor, ylim=c(0,15000))
```

Both mean and median show that male workers earn more compared to female workers.

So let us see if female workers actually work less hours compared to male workers.

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'



#function for Latex table output since it is repeated often
latex_table <- function(data, title = "Table", annotation= NULL) {
  # Create gt table
  tab <- data %>%
    gt() %>%
    tab_header(title = title)
  
  # Convert to LaTeX and set float position to [H]
  latex_code <- as.character(as_latex(tab))
  latex_code <- gsub("\\\\begin\\{table\\}", "\\\\begin{table}[H]", latex_code)
  
  # Output LaTeX code
  cat(latex_code)
}

```

### Means for extra hours depending on gender

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'



#---------------------------------------------------------

#means for extra hours depending on gender

# Summarize mean of xtrahrs by gender
tab<-tapply(labor$xtrahrs, labor$gender, mean, na.rm = TRUE)
summary_labor<-as.data.frame(tab)
colnames(summary_labor)<-c("Mean of extra hours")

summary_labor <- cbind(gender = rownames(summary_labor), summary_labor)

rownames(summary_labor) <- NULL

latex_table(summary_labor, "Mean of extra hours by gender")

```

### Full-time and part-time workers for each gender

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: asis

#full-time vs part-time for each gender



tab <- table(labor$gender, labor$ftptmain)

# Convert to data frame keeping wide structure
summary_labor <- as.data.frame.matrix(tab)

# Optionally add row names as a proper column
summary_labor <- cbind(gender = rownames(summary_labor), summary_labor)
rownames(summary_labor) <- NULL


latex_table(summary_labor, "Full-time anf part-time workers by gender")

```


### Reasons for part-time work by gender

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: 'asis'


#full-time vs part-time for each gender

tab <- table(labor$gender, labor$ftptmain)

# Convert to data frame keeping wide structure
summary_labor <- as.data.frame.matrix(tab)

# Optionally add row names as a proper column
summary_labor <- cbind(gender = rownames(summary_labor), summary_labor)
rownames(summary_labor) <- NULL


latex_table(summary_labor, "Full-time anf part-time workers by gender")

# reasons for part-time vs gender


tab<-xtabs(~ whypt+gender, data=labor )

summary_labor <- as.data.frame.matrix(tab)

summary_labor$Reason <- whypt_labels  # 


latex_table(summary_labor, "Reasons for part-time work by gender")





#Conclusion : women stay away from work taking care of the children approx 10x then men
```

<!-- We nay immediately notice conclusion: women stay away from work taking care of the children approx 10x then men -->

```{r}
#| eval: false
#| message: false
#| warning: false
#| include: false
#| results: asis
#hour rate* usual hours=income  vs type of employment
#tapply(0.01*labor$hrlyearn, labor$PERMTEMP, mean, na.rm = TRUE)
#tapply(0.01*labor$hrlyearn, labor$PERMTEMP , median, na.rm = TRUE)
# Permanent or temporary job (field PERMTEMP)
# 1	Permanent
# 2	Temporary, seasonal job
# 3	Temporary, term or contract job
# 4	Temporary, casual or other temporary jobs


```

# Unemployment

### Unemployment vs gender (non-weighted and weighted)

```{r}
#| echo: false
#| warning: false
#| results: asis

########################################################
#absolute numbers of male and female workers 

tab<-xtabs(~lfsstat + gender , data=labor)

summary_labor <- as.data.frame.matrix(tab)

# Add row names as a proper column
summary_labor <- cbind(Status = rownames(summary_labor), summary_labor)
rownames(summary_labor) <- NULL

latex_table(summary_labor, "Quantities of male and female workers")

########################################################
# unemployment rates vs gender 
tab<-prop.table(tab, margin = 2)
summary_labor <- as.data.frame.matrix(tab)

# Add row names as a proper column
summary_labor <- cbind(Status = rownames(summary_labor), summary_labor)
rownames(summary_labor) <- NULL

latex_table(summary_labor,"Nonweighted unemployment rates by gender")

########################################################
# unemployment rates vs gender among those in labour force
tab<-prop.table(tab[1:3,], margin = 2)

summary_labor <- as.data.frame.matrix(tab)

# Add row names as a proper column
summary_labor <- cbind(Status = rownames(summary_labor), summary_labor)
rownames(summary_labor) <- NULL

latex_table(summary_labor,"Nonweighted unemployment rates by gender")


########################################################
#Weighted unemployed rates

source("bootstrapStatCa.R")
#print(unemprate_by_prov(final_bs))

# Calculating unemployed rates by province, gender, and immigrant status using  bootstrap

#print(unemprate_by_gender(final_bs))
unemp_table <- unemprate_by_gender(final_bs)
unemp_table<-as.data.frame(t(unemp_table))
unemp_table$var<-rownames(unemp_table)
unemp_table <- unemp_table[, c("var", setdiff(names(unemp_table), "var"))]

colnames(unemp_table) <- as.character(unlist(unemp_table[1, ]))
unemp_table <- unemp_table[-1, ]


latex_table(unemp_table,"Weighted unemployment rate by gender")

```

### Unemployment vs immigrant status (non-weighted and weighted)

```{r}
#| echo: false
#| warning: false
#| results: asis


tab<-xtabs(~lfsstat + immig , data=labor)
#absolute numbers of immigrant and non-immigrant workers 

summary_labor <- as.data.frame.matrix(tab)

# Add row names as a proper column
summary_labor <- cbind(status = rownames(summary_labor), summary_labor)
rownames(summary_labor) <- NULL

latex_table(summary_labor,"Quantities of immigrant and non-immigrant workers")


# unemployment rates vs immigrant status 

summary_labor <- as.data.frame.matrix(prop.table(tab, margin = 2))

# Add row names as a proper column
summary_labor <- cbind(status = rownames(summary_labor), summary_labor)
rownames(summary_labor) <- NULL

latex_table(summary_labor,"Nonweighted unemployment rates by immigrant status")




# Calculating unemployed rates by immigrant status using  bootstrap

#print(unemprate_by_immig(final_bs))
unemp_table <- unemprate_by_immig(final_bs)
#unemp_table <- cbind(status = rownames(unemp_table), unemp_table)

unemp_table<-as.data.frame(t(unemp_table))
unemp_table$var<-rownames(unemp_table)
unemp_table <- unemp_table[, c("var", setdiff(names(unemp_table), "var"))]

colnames(unemp_table) <- as.character(unlist(unemp_table[1, ]))
unemp_table <- unemp_table[-1, ]
latex_table(unemp_table,"Unemployment Rate by Immigrant Status")


```

# Wages and Demographics

Family matter affects the hours and wages of female workers. Let us see how wages depend on age and gender.

## Wages vs age (using linear regression and poly approximation)

First we check how wages depend on age of workers of both genders.

```{r}
#| echo: false
#| message: false
#| warning: false

#tapply(labor$hrlyearn, labor$age_12 , mean, na.rm = TRUE)
tmp <- tapply(labor$hrlyearn, labor$age_12, mean, na.rm = TRUE)

#with(labor, tapply(0.01 * hrlyearn * utothrs, age_12, mean, na.rm = TRUE))

# Create data frame; age_12 numeric for modeling, age_group factor for labels
df <- data.frame(
  age_12 = as.numeric(names(tmp)),        # numeric age codes (1,2,3...)
  salary = as.numeric(tmp)
)
df$age_group <- factor(df$age_12, levels = as.numeric(names(age_12_labels)), labels = age_12_labels)

# Fit polynomial regression on numeric age_12
model <- lm(salary ~ poly(age_12, 2), data = na.omit(df))
df$pred <- predict(model)

ggplot(df, aes(x = age_group, y = salary, color = "black")) +
  geom_point(size = 2, show.legend = FALSE) +    # no legend for points
  geom_line(aes(y = pred, group = 1), color = "blue", size = 0.5) +
  labs(title = "Wages vs age", x = "age group", y = "Mean Wages") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

```

## Wages vs age (using linear and quadratic polynоmial regressions)

```{r}
#| echo: false
#| message: false
#| warning: false

# LINEAR REGRESSION for SALARY vs age (age_12 is treated as numeric)
model <- lm(0.01 * hrlyearn * utothrs ~ as.numeric(age_12), data = labor)
#model

# SECOND ORDER POLYNOMIAL REGRESSION
#lm(formula = 0.01 * hrlyearn * utothrs ~ poly(as.numeric(age_12), 2), data = labor)

# Scatter plot with linear and polynomial regression lines
p1 <- labor %>%
  ggplot(aes(x = as.numeric(age_12), y = 0.01 * hrlyearn * utothrs)) +
  geom_point(color = "orange") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "blue") +
  scale_x_continuous(
    breaks = 1:length(age_12_labels),
    labels = age_12_labels
  ) +
  labs(x = "age", y = "wages means: actual hours") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))


# Boxplot by age_12 factor
p2 <- labor %>%
  ggplot(aes(x = age_12, y = 0.01 * hrlyearn * utothrs)) +
  geom_boxplot(fill = "orange") +
  scale_x_discrete(labels = age_12_labels) +  # <-- show long labels for age
  labs(x = "age", y = "wages means: actual hours") +
  theme_bw()+  
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

# Display plots side by side
p1 

p2

```

## Wages vs age for males and females

```{r}
#| echo: false
#| message: false
#| warning: false


# plot shows salary (based on usual hours and usual wages) by gender
pl1<-labor %>%
  ggplot(aes(x = age_12, y = 0.01 * hrlyearn * utothrs, color = gender)) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun = mean, geom = "line", aes(group = gender), color="grey", linewidth = 0.3) +
    scale_x_discrete(labels = age_12_labels) +  # <-- show long labels
  labs(x = "age", y = "wages means: usual hours") +
  theme_bw()+  
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

```

```{r}
#| echo: false
#| message: false
#| warning: false

#hyst plot showing paid plus unpaid extra hours for Females and Males during the week of sampling


pl2<-labor %>%
  ggplot(aes(x = age_12, y = 0.01 * hrlyearn *atothrs, color = gender)) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun = mean, geom = "line", aes(group = gender), color = "grey", linewidth = 0.3) +
  scale_x_discrete(labels = age_12_labels) +  # <-- show long labels for age
  labs(x = "age", y = "wages means: actual hours") +
  theme_bw()+  
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
pl1 
pl2


```

## Wages depend on immigrant status

```{r}
#| echo: false
#| message: false
#| warning: false

#hour rate* usual hours=income  vs immigrant status

tapply(0.01*labor$hrlyearn*labor$utothrs, labor$immig , mean, na.rm = TRUE)
tapply(0.01*labor$hrlyearn*labor$utothrs, labor$immig , summary, na.rm = TRUE)
# boxplot(0.01*hrlyearn*utothrs ~ immig ,data=labor)
# boxplot(hrlyearn ~ immig ,data=labor)
# boxplot(utothrs ~ immig ,data=labor)

ggplot(data=labor, aes(x=immig,y=0.01*hrlyearn*utothrs))+
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75))+
  theme_bw() +
  labs(x="immigration status",y="Salary")
```

## Wages vs tenure (using linear regression)

```{r}
#| echo: false
#| message: false
#| warning: false

#idea: income=hours*rate for different ages, non-linear dependence vs linear regression

#LINEAR REGRESSION for SALARY vs tenure
model<-lm(formula = 0.01*hrlyearn*utothrs~tenure,data=labor)

labor %>% ggplot(aes(x=tenure,y=0.01*hrlyearn*utothrs,color = gender))+
geom_point( size = 0.3)+
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "tenure in monthes", y = "Salary") +
  theme_minimal()


```

# Model-based analysis: What factors are crucial for wages

## Wages clustering : using UMAP and K-means

```{r}
#| echo: false
#| message: false
#| warning: false

library(uwot)
#library(umap) #clustering
library(knitr)




tmp.data <- labor[, c("hrlyearn","paidot","unpaidot",
        "xtrahrs", "noc_43", "noc_10","naics_21",
        "utothrs","atothrs","hrsaway","tenure","age_12")]  #"uhrsmain","ahrsmain",

cols <- c("noc_43", "noc_10", "naics_21", "age_12")

tmp.data[cols] <- lapply(tmp.data[cols], function(x) as.numeric(replace(x, x == "", 0)))

keep_rows <- complete.cases(tmp.data)


tmp.data <- tmp.data[keep_rows, ]

clusters <- kmeans(tmp.data, centers = 3)$cluster


umap_result<-umap(as.matrix(tmp.data))

# Convert umap layout to data frame
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")

# try colors defined by K-means clusterization
# Combine with original data 
combined <- cbind(umap_df, tmp.data)

# Correlations: how much each variable contributes to the axis UMAP1, UMAP2

cor_matrix <- cor(combined[, c("UMAP1", "UMAP2")],
                  combined[, -c(1,2)],
                  use = "complete.obs")

# Convert to data frame and format
cor_df <- as.data.frame(t(cor_matrix))

# Render nicely
kable(cor_df, digits = 2, caption = "Correlation of UMAP dimensions with other variables")


#clusters <- kmeans(umap_result, centers = 3)$cluster

umap_df$col <- as.factor(clusters)

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = col)) +
  geom_point(size = 2, alpha = 0.7) +
  theme_bw() +
  labs(title = "UMAP Clustering", x = "UMAP 1", y = "UMAP 2", color = "Cluster")

# Factor column for coloring 
tmp.labels <- factor(ifelse(0.01 * tmp.data$hrlyearn * tmp.data$utothrs > 10500, "Yes", "No"))

print("Factor: wage based on utothrs > 10500 ")
#tmp.labels <- labor[1:3000,][keep_rows,"gender"]

umap_df$col <- tmp.labels 

# Plot with ggplot2
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = col)) +
  geom_point(size = 1.5) +
  labs(title = "UMAP Projection", x = "UMAP 1", y = "UMAP 2", color = "Factor") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))


```

## Clustering using DBSCAN

```{r}
#| echo: false
#| message: false
#| warning: false

library(mclust)

tmp.data <- labor[, c("hrlyearn","paidot","unpaidot",
        "xtrahrs", "noc_43", "noc_10","naics_21",
        "utothrs","atothrs","hrsaway","tenure","age_12")]  #"uhrsmain","ahrsmain",

cols <- c("noc_43", "noc_10", "naics_21", "age_12")
tmp.data[cols] <- lapply(tmp.data[cols], function(x) as.numeric(replace(x, x == "", 0)))

keep_rows<-complete.cases(tmp.data)

tmp.data <- tmp.data[keep_rows, ]

mclust_result <- Mclust(tmp.data)

# summary and method  Mclust if needed
#table(mclust_result$classification) 

# summary(mclust_result)
# mclust_result$G  # Number of components selected
# mclust_result$modelName 

clusters <- mclust_result$classification

#visualization using umap
umap_result<-umap(as.matrix(tmp.data))

# Convert umap layout to data frame
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")

umap_df$col <- as.factor(clusters)

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = col)) +
  geom_point(size = 2, alpha = 0.7) +
  theme_bw() +
  labs(title = "UMAP Clustering", x = "UMAP 1", y = "UMAP 2", color = "Cluster")

# only one cluster -> statistical conclusion that no meaningful subgrouping was found
```

## Linear regression

Wages are determined by the hourly rate (**hrlyearn**) and the typical number of hours worked (or the actual total hours - fields **utothrs, atothrs paidot, unpaidot**). Both hourly rate and hours worked exhibit a higher correlation with the UMAP embedding coordinates, suggesting they contribute significantly to the underlying variance structure of the dataset.

In the next step we check what factors define the hourly rate (**hrlyearn**) and usual hours (**utothrs**).

### Model on the hourly rate

```{r}
#| echo: false
#| message: false
#| warning: false

#rows with no NAs
complete_labor <- labor[complete.cases(labor[,c("utothrs","age_12","educ", "gender","cowmain","marstat","agyownk","naics_21","noc_10","noc_43","prov") ]), ]

#we consider age a numeric variable
complete_labor$age_12<-as.numeric(complete_labor$age_12)

#null_model <- lm(hrlyearn ~ 1, data = complete_labor)

#linear regression model  and the summary
model <- lm(hrlyearn ~ utothrs+gender+educ +cowmain+agyownk+age_12+noc_10+prov , data = complete_labor)

summary(model)
```

### Model on usual total working hours

```{r}
#| echo: false
#| message: false
#| warning: false


### Model on utothrs - Usual hours worked per week at all jobs
## or should it be atothrs - Actual hours worked per week at all jobs



alt_model <- lm(atothrs ~ gender+educ +cowmain+agyownk+naics_21+prov , data = complete_labor)

summary(alt_model)

#AIC and BIC testing what model is more preferable
AIC(model)
BIC(model)

#nested model compare
anova(model, alt_model)

```

#Chi-Squared and ANOVA Tests Dependence Between Categorical Variables

```{r}
#| echo: false
#| message: false
#| warning: false






#correlation between numeric parameters to see their contribution to the lin model

chisq.test(table(labor$age_12, labor$marstat))
chisq.test(table(labor$agyownk, labor$marstat))
chisq.test(table(labor$agyownk, labor$age_12))
chisq.test(table(labor$naics_21, labor$noc_10))

chisq.test(table(labor$naics_21, labor$noc_43))

library(DescTools)

EtaSq(aov(hrlyearn ~ prov+gender, data = labor))  # Strength of association

cor(labor$utothrs, labor$hrlyearn, use = "complete.obs")


#step_model <- step(null_model, scope = list(lower = null_model, upper = full_model), direction = "both",na.action = na.omit)
```

<!-- Bootstrap: non-weighted and weighted -->

```{r}
#| eval: false
#| message: false
#| warning: false
#| include: false

# Confidence intervals via non-weighted bootstrap  
# seed <- 1234
# set.seed(seed_value)

  
mean.function <- function(x, index) {
    d <- x[index]     # This first line will go in ever bootstrap function you make.
    return(mean(d,na.rm = TRUE))  
  }

BootHR<-boot(data = labor$hrlyearn, statistic = mean.function, R=1000)
head(BootHR$t)
BootHR.graph <- data.frame(xbar=BootHR$t)

ggplot(BootHR.graph, aes(x=0.01*xbar)) +
  geom_histogram(color="darkblue", fill="lightblue") + 
  ggtitle('Estimated Sampling distribution of Means of Hour Rate (hrlyearn)' )

# 0.95 confidence interval
quantile( BootHR$t, probs=c(.025, .975) ) 
boot.ci(BootHR, type = "perc", conf = 0.95)  
```

```{r}
#| eval: false
#| include: false
#| echo: false
#| message: false
#| warning: false

################################################################################
#NOT USED
################################################################################

# Part-time workers study



# PART TIME vs gender

# reasons for part-time vs gender
tab<-xtabs(~ gender + whypt, data=labor )
tab
#Conclusion : women stay away from work taking care of the children approx 10x then men

```

```{r}
#| eval: false
#| include: false

#| echo: false
#| message: false
#| warning: false

df_tab <- as.data.frame(tab)

df_tab %>%
 # filter(whypt %in% c("2","3","6")) %>%
  ggplot( aes(x = whypt , y=Freq, color=gender)) +
  geom_point(size=3)+
  theme(legend.position="top")+
  labs(x = "Why part-time job", 
       caption = "0	- Other reasons
1 -	Own illness or disability
2	- Caring for children
3	- Other personal or family responsibilities
4	- Going to school
5	- Personal preference
6	- Business conditions or could not find full-time work, looked for full-time work in last month
7	- Business conditions or could not find full-time work, did not look for full-time work in last month
", 
       y = "Number")+
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0, size = 10)
  )
```

```{r}
#| eval: false
#| include: false

#| echo: false
#| message: false
#| warning: false

# gender vs REASONS FOR LEAVING JOB
tab<-xtabs(~ gender + whyleftn, data=labor )
tab
df_tab <- as.data.frame(tab)

df_tab %>%
  filter(whyleftn %in% c("1","2","3","4","5")) %>%
  ggplot( aes(x = whyleftn , y=Freq, color=gender)) +
  geom_point(size=3)+
  theme_bw()+
  theme(legend.position="top")+
  labs(x = "Why left job in the last 12 months", 
       caption="
02	Job leavers, caring for children
03	Job leavers, pregnancy
04	Job leavers, personal or family responsibilities
05	Job leavers, going to school",
       y = "Number")+
theme(
  plot.caption = element_text(hjust = 0, size = 10)
)

# Conclusion: there are more males leaving work for school then females 
#
# part time going to school males 1664, females 2349
# left work because of the going to school males 810, females 738
# possible trend: males less likely work full or part-time during attending school
#compared to females


```
